SDSS-IV MaNGA: The Radial Profile of Enhanced Star Formation in Close Galaxy Pairs

### 2020-10-21 maintex.tex
further edits up to page 5

We identify potential paired galaxies by overlaying SDSS photometric objects over the MaNGA fields of view. We manually inspect each MaNGA field, removing photometric objects which are over-deblended galaxy fragments and adding any objects missed by the SDSS photometric catalog. At this stage the object catalog includes any foreground stars and foreground/background galaxies along with the potential paired galaxies.

@@

From the 6573 galaxies with good spectra, 404 of the MaNGA IFUs have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in Section \ref{sec:data}. These requirements reduce our sample down to 87 star forming galaxies (SFGs).

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. This give use a total of 175 MaNGA targets in our pair sample.

@@

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep{Noeske:2007}. We will compare our interacting galaxies with isolated galaxies of a similar stellar mass and redshift, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

In the first method, we only control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bins. While the redshift in not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

In the second method, we wish to account for both stellar mass and redshift. To do this we select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within a 0.1 dex stellar mass limit and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. Since the MaNGA sample has a tight distribution in stellar mass and redshift, the stellar mass and MaNGA subsample restriction effectively sets a $z$ $\lesssim$ $\sim$0.025 redshift cut.

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we will down-select the total number of acquired control galaxies down to 20 controls randomly. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. Only 5 of the paired galaxies required an extra iteration to find 20 controls.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Radial Profiles of Star Formation}\label{sec:analysis}

@@

%%% SPFIT
We use our spectral fitting code, {\sc spfit}, to get measurements of emission line fluxes, equivalent widths, velocities, and velocity dispersions of 19 emission lines and measurements of stellar masses, ages, [Fe/H], and the kinematics of the stellar populations.

@@

We measure the SFR from the extinction corrected H$\alpha$ luminosity, $L_{{\rm H}\alpha}$.  We use the SFR formula, Equation \ref{eq:sfr}, from \citet{Murphy:2011} which uses a Kroupa IMF, Solar metallicity, a constant SFR at an age of 100 Myr, and Case-B recombination.

\begin{equation}\label{eq:sfr}
\frac{\rm{SFR}}{M_{\odot} \, \rm{yr^{-1}}} = \frac{L_{\rm H\alpha}}{1.86 \times 10^{41}\, \rm{erg\,s}^{-1}}
\end{equation}

Since the stellar mass of a galaxy is not uniformly distributed within the galaxy, we normalize the SFR by the local stellar mass, $M$, giving us the specific star formation rate (sSFR). The local stellar masses used here is the stellar mass calculated by {\sc SPFIT}'s fit to the stellar continuum within each MaNGA spaxel (spatial pixel).

\begin{equation}
{\rm sSFR} = \frac{\rm SFR}{M}
\end{equation}

This will allow us to compare regions of high mass surface density to regions of low mass surface density.

@@

We fit the geometry with apertures from both catalogs when available. When using the first method of pair-control comparison, where pairs and controls are grouped into stellar mass bins, we fit the control galaxies with the NSA apertures. When making the comparison with the second method, paired galaxies fitted with the NSA aperture are compared against controls fitted with the NSA aperture and paired galaxies fitted with the \citet{Simard:2011} apertures are compared against controls fitted with the \citet{Simard:2011} apertures. Further, when defining the a galaxy's geometry with the \citet{Simard:2011} apertures we also use the masses given in the catalog for the total stellar mass of the galaxy.

@@

Once the geometry of the galaxies are set, we can build the radial profiles. The spaxels are binned into radius increments of 0.2 \reff\ from 0.0$-$2.6 \reff. Within each radius bin we take the median of the specific star formation rate. The profiles' errors are the standard error of the mean of the data within each radius bin. We create one of these individual averaged radial profiles for every MaNGA galaxy.

@@

In the first method, paired and control galaxies are grouped into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5. Within each stellar mass bin, we create a median profile of log(sSFR) as a function of galactocentric radius for both paired and control galaxies. The control profiles are constructed using all available control galaxies within each mass bin. The error associated with the profile is the standard error of mean of the data at each radius bin. We show these ``stacked" profiles for control galaxies in the left hand panel of Figure \ref{fig:ssfr_prof} and the stacked profiles for paired galaxies in the middle panel of Figure \ref{fig:ssfr_prof}.

The radial extent of our profiles go out to 2.6 \reff; however,  67\% of the MaNGA sample is designed to cover only 1.5 \reff\ of the galaxy. Also, the MaNGA galaxies are not always centered on a spaxel center, so the 0$^{\rm th}$ spaxel may fall outside of the innermost radii bin. Because of these issues, the centers or the outskirts of the stacked profiles may only represent a small subset of the galaxies in that mass bin. To account for this, we ``trim" the stacked profiles where a radius bin contains data for less than 50\% of the total set of galaxies. We represent the trimmed regions in our figures with dashed lines. In Figure \ref{fig:ssfr_prof}, galaxies below \logm\ $\le$ 10.0 have their innermost radii bin trimmed due to insufficient coverage.

Once the stacked profiles are made, we take the difference between the stacked profiles of the paired galaxies and the control galaxies, pair - control, in log space (this means that the difference profiles really represent a ratio between the pairs and controls in linear space). This gives us difference profile, $\Delta$log(sSFR), which shows us where the sSFR is enhanced or suppressed (shown in the right hand panel of Figure \ref{fig:ssfr_prof}).

The control galaxies are quenched in their centers with respect to their disks and the difference between the sSFR in their centers and their disks increases with higher stellar masses (which was previously seen in the MaNGA survey in \citet{Belfiore:2018}). The paired galaxies show flattened star formation profiles where the level of the sSFR remains roughly consistent across their disks except for paired galaxies with stellar masses above \logm\ $=$ 10.5 which still show some central quenching.

When taking the difference between the pairs and controls, the paired profiles are centrally enhanced by $\sim$0.35 $\pm$ 0.1 dex which then falls to zero around 1.2 \reff. In the outskirts of the pairs beyond 1.2 \reff, the pairs feature lightly suppressed sSFR of 0.0$-$0.1 dex. All of the mass ranges except for the highest mass range, \logm\ $=$ 11.0$-$11.5, trend closely to the median profile. For the most massive galaxies, \logm\ $=$ 11.0$-$11.5, the enhancement to the sSFR within 0.5 \reff\ is higher than the median profile at 0.5$-$0.6 dex. The sSFR enhancement in the sample seems to be mass independent except for the most massive galaxies which feature higher levels of sSFR enhancement.

It is also relevant to note that the driving factor behind the central enhancement to the sSFR in the difference profiles is really the quenching in the centers of the controls. While the mechanism for the quenching in galaxies is in dispute, the merger process seems to be capable of rejuvenating star formation in the centers of these galaxies resulting in the flatter sSFR profiles seen in the paired galaxies. Further, since the sSFR enhancement in the difference profiles is largely driven by the quenched controls, it may be more appropriate to describe isolated galaxies as being centrally quenched in comparison to paired galaxies or that the profiles of paired galaxies are flattened in their centers with respect to their controls.

>>>>

To identify potential paired galaxies covered by individual IFUs, we start by overlaying SDSS photometric objects over each MaNGA fields of view. We manually inspect each MaNGA field, removing photometric objects which are over-deblended galaxy fragments and, very rarely, adding any objects missed in the SDSS photometric catalog. At this stage the object catalog includes foreground stars and foreground/background galaxies along with the potential paired galaxies.

@@

Among 6142 MaNGA IFUs considered here, 404 have multiple galaxy objects within the IFU. We further restrict the sample by setting a relative velocity cut of $\Delta v$ $<$ 500 km s$^{-1}$ to remove projected companions. Given redshift range of the MaNGA sample and the size of MaNGA's IFUs, the maximum projected separation for a companion galaxy in the IFU is $\sim$40 kpc. Again, the galaxy sample is also restricted to galaxies with a stellar mass range of \logm\ $=$ 9.0$-$11.5 and galaxies in the Primary or Secondary MaNGA subsamples. We also require that the galaxies are classified as star forming as described in Section \ref{sec:data}. These requirements reduce our sample down to 87 star forming pairs.

@@

From the NSA catalog's 641,409 galaxies, we find 492 galaxies which are paired to MaNGA targets. After restricting the sample to SFGs, we have another 133 MaNGA targets with paired galaxies outside of the IFU. MaNGA targets which have both an inside-IFU and an outside-IFU pair are left to the inside-IFU sample. This give us a total of 175 close pairs.

@@

It has been shown that the SFR in galaxies is dependent on both the stellar mass and the redshift of the galaxies \citep[e.g.,][]{Noeske:2007}. We will compare our interacting galaxies with isolated galaxies of a similar stellar mass and redshift, to account for these other dependencies. To do this, we will use two different methods of pair - control comparison.

In the first method, we only control for the stellar mass between the pairs and controls. We split both the pair and control samples into five evenly spaced stellar mass bins over the range, \logm\ $=$ 9.0$-$11.5. The pairs are then compared to their respective controls within each stellar mass bins. While the redshift is not constrained in this method, we will show in Section \ref{sec:mass-bin} that this method is sufficient to reveal the merger induced star formation.

In the second method, we wish to account for both stellar mass and redshift. To do this we select a subsample of 20 control galaxies for each paired galaxy. We first select all of the control galaxies which are within a 0.1~dex stellar mass limit and within the same MaNGA subsample (e.g. Primary or Secondary) as the paired galaxy. Since the MaNGA sample has a tight distribution in stellar mass and redshift, the stellar mass and MaNGA subsample restriction effectively sets a $\Delta z$ $\lesssim$ $\sim$0.025 redshift tolerance.

Setting the above requirements, most paired galaxies will find between 20$-$100 control galaxies. Since we want each paired galaxy to be treated in a similar manner, we randomly down-select the total number of acquired control galaxies to 20. A given control galaxy may be selected for multiple pairs by the pipeline; most of the control galaxies are used at least once and the average number of times a given control is reused is between 4$-$6 times. In the cases where a paired galaxy does not initially acquire 20 control galaxies we iteratively expand the stellar mass limit by 0.1 dex until at least 20 control galaxies are found. Only 5 of the paired galaxies required an extra iteration to find 20 controls.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis Methods}\label{sec:analysis}

@@

%%% SPFIT
For each MaNGA 1\arcsec$\times$1\arcsec\ spaxel, we use our spectral fitting code, {\sc spfit}, to get measurements of emission line fluxes, equivalent widths, velocities, and velocity dispersions of 19 emission lines and measurements of stellar masses, ages, [Fe/H], and the kinematics of the stellar populations.

@@

We measure the SFR from the extinction corrected H$\alpha$ luminosity, $L_{{\rm H}\alpha}$.  We use the SFR formula, Equation \ref{eq:sfr}, from \citet{Murphy:2011} which uses a Kroupa IMF, Solar metallicity, a constant SFR at an age of 100 Myr, and Case-B recombination: 
\begin{equation}\label{eq:sfr}
\frac{\rm{SFR}}{M_{\odot} \, \rm{yr^{-1}}} = \frac{L_{\rm H\alpha}}{1.86 \times 10^{41}\, \rm{erg\,s}^{-1}}.
\end{equation}

Since the stellar mass of a galaxy is not uniformly distributed within the galaxy, we normalize the SFR by the stellar mass in the same spaxel, giving us the specific star formation rate (sSFR):
\begin{equation}
{\rm sSFR} = \frac{\rm SFR}{M}
\end{equation}
The local stellar masses used here is derived from {\sc SPFIT}'s best-fit stellar continuum. Utilizing sSFR, instead of SFR, allows us to compare regions of high mass surface density to regions of low mass surface density.

@@

We fit the geometry with apertures from both catalogs when available. When using the first method of pair-control comparison, where pairs and controls are grouped into stellar mass bins, we fit the control galaxies with the NSA apertures. When making the comparison with the second method, paired galaxies fitted with the NSA aperture are compared against controls fitted with the NSA aperture and paired galaxies fitted with the \citet{Simard:2011} apertures are compared against controls fitted with the \citet{Simard:2011} apertures. Further, when defining a galaxy's geometry with the \citet{Simard:2011} apertures we also use the masses given in the catalog for the total stellar mass of the galaxy. Note that although we carefully treat the two catalogs separately whenever possible, we find excellent agreement in the geometry parameters of isolated galaxies in both catalogs.

@@

Once the geometry of the galaxies are set, we can build azimuthally averaged radial profiles. The spaxels are binned into radius increments of 0.2 \reff\ from 0.0$-$2.6 \reff. Within each radius bin we take the median of the specific star formation rate. The profiles' errors are the standard error of the mean of the data within each radius bin. We build an azimuthally averaged radial profile for each MaNGA star forming galaxy and later stack and differentiate these profiles in the subsequent analysis.

@@

In the first method, paired and control galaxies are grouped into five evenly spaced stellar mass bins between \logm\ $=$ 9.0$-$11.5. Within each stellar mass bin, we create a median profile of log(sSFR) as a function of \reff-scaled galactocentric distance for both paired and control galaxies. The control profiles are constructed using all available control galaxies within each mass bin. The error associated with the ``stacked'' profile is the standard error of mean of the data at each radius bin. We show these stacked profiles for control galaxies in the left hand panel of Figure \ref{fig:ssfr_prof} and the stacked profiles for paired galaxies in the middle panel of Figure \ref{fig:ssfr_prof}. {\bf QUESTION: for the inside-IFU pairs, did you use only the primaries to build these stacked profiles? If so, this needs to made clear in the text.} Star formation in the control galaxies are quenched in their centers with respect to their disks and the difference between the sSFR in their centers and their disks increases with stellar mass, consistent with previously published results using the MaNGA survey \citep{Belfiore:2018}. In contrast, the paired galaxies show flatter sSFR profiles where the level of the sSFR remains roughly consistent across their disks except for paired galaxies with stellar masses above \logm\ $=$ 10.5, which still show some central quenching.

{\bf COMMENT: I am not sure the trimmed regions could be so limited. If this is indeed correct, then we don't need to spend a paragraph on something so negligible and not to highlight the trimed regions in the plots.} 
The radial extent of our profiles reaches 2.6 \reff; however, 67\% of the MaNGA sample (i.e., the Primary subsample) is designed to cover only 1.5 \reff\ of the galaxy. Also, the MaNGA galaxies are not always centered on a spaxel center, so the 0$^{\rm th}$ spaxel may fall outside of the innermost radii bin. Because of these issues, the centers or the outskirts of the stacked profiles may only represent a small subset of the galaxies in that mass bin. To account for this, we ``trim" the stacked profiles where a radius bin contains data for less than 50\% of the total set of galaxies. We represent the trimmed regions in our figures with dashed lines. In Figure \ref{fig:ssfr_prof}, galaxies below \logm\ $\le$ 10.0 have their innermost radii bin trimmed due to insufficient coverage.

We then take the difference between the stacked profiles of the paired galaxies and the control galaxies, pair - control, in log space (this means that the difference profiles really represent a ratio between the pairs and controls in linear space). This gives us difference profile, $\Delta$log(sSFR), which shows us where the sSFR is enhanced or suppressed (shown in the right hand panel of Figure \ref{fig:ssfr_prof}). Across all stellar mass bins, the sSFR of paired galaxies are centrally enhanced by $\sim$0.35$\pm$0.1~dex, which gradually falls to zero around $\sim$1.2 \reff. In the outskirts of the pairs beyond 1.2 \reff, the pairs feature lightly suppressed sSFR of 0.0$-$0.1 dex. All of the mass ranges except for the highest mass range, \logm\ $=$ 11.0$-$11.5, trend closely to the median profile. For the most massive galaxies, \logm\ $=$ 11.0$-$11.5, the enhancement to the sSFR within 0.5 \reff\ is higher than the median profile at 0.5$-$0.6 dex. The sSFR enhancement in the sample seems to be mass independent except for the most massive galaxies which feature higher levels of sSFR enhancement.

Alternatively, the central enhancement of the sSFR in the paired galaxies can be interpreted as the quenching of star formation in the centers of isolated galaxies. While the mechanism for quenching in galaxies remains under debate, the merger process seems to be capable of rejuvenating star formation in the centers of these galaxies, resulting in the flatter sSFR profiles. 
%Further, since the sSFR enhancement in the difference profiles is largely driven by the quenched controls, it may be more appropriate to describe isolated galaxies as being centrally quenched in comparison to paired galaxies or that the profiles of paired galaxies are flattened in their centers with respect to their controls.
